The algebraic hierarchy in type theory using type classes

### 2010-05-07 jar.tex
Small changes

This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from \lstinline|Type| to \lstinline|Prop| reflects the fact that \lstinline|ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opaque definitions. The former can be unfolded,
whereas this is impossible for the latter. Since we never need to distinguish between two proofs of
a propositions, Propositions can be Opaque.}
we will return to this issue in Section~\ref{PI}.
We can now define \lstinline|EquivalenceRelation| without awkward equalities:
\begin{lstlisting}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: relation er_carrier
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}

@@

So far, we have only looked at the definitions of the structures themselves, and have not yet
considered the perspective of someone trying to prove some theorems about these abstract structures.
We would like to approximate the common mathematical vernacular, where one
simply says:

@@

We choose to use \lstinline|&| here and reserve the notation \lstinline|*| for ring multiplication.

We use = for setoid equality, since Leibniz equality seems less important for the formalization of
mathematics. More precisely, setoid equality is the basic concept. In special case, one can prove
that setoid equality coincides with Leibniz equality.

We would have like to use a haskell style definition
\begin{lstlisting}
Class Equiv A := (=): relation A.
\end{lstlisting}

Unfortunately, currently this is not supported in Coq.

@@

It is important to point out that although the expression now \emph{looks} like it is bound to some
specific semigroup structure (which with traditional approaches would imply projections from bundles
and/or reference to proofs), it is really only referring near-directly to the actual operations
involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
may use in the proof. This lack of projections keeps our terms small and independent, and keeps
rewriting simple and sane; see Section~\ref{canonical}.

@@

where e.g.\ \lstinline|RingPlus| is the operation class for the semigroup operator.
The notation \lstinline|:>| makes \lstinline|semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
provide a ring.

@@

First, multiple inheritance becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.

@@

Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue. Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a |SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.

Fourth, \emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
can find the equivalence relation in the context.\footnote{
A similar style should be possible for, say, the |Ring| tactic, instead of
declaring the ring structure by a separate command, we would rely on type class resolution to find it.}
We note that \lstinline|op| does not depend on the proof that \lstinline|e| is an equivalence. As explained above we use Coq's implicit quantification (\lstinline|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a theorem. Both features are new in Coq and stem from the type class implementation.

>>>>

This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from \lstinline|Type| to \lstinline|Prop| reflects the fact that \lstinline|ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opaque definitions. The former can be unfolded,
whereas this is impossible for the latter. Since we never need to distinguish between two proofs of
a propositions, propositions can be opaque.}
we will return to this issue in Section~\ref{PI}.
We can now define \lstinline|EquivalenceRelation| without awkward equalities:
\begin{lstlisting}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: relation er_carrier
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}

@@

So far, we have only looked at the definitions of the structures themselves, and have not yet
considered their use when proving theorems about these structures.
We would like to approximate the common mathematical vernacular, where one
simply says:

@@

We choose to use \lstinline|&| here and reserve the notation \lstinline|*| for ring multiplication. We use = for setoid equality, since Leibniz equality seems less important for the formalization of
mathematics. More precisely, setoid equality is the basic concept. In some special cases, one can prove
that setoid equality coincides with Leibniz equality.
%Listings don't seem to work will with footnotes.
We would have like to use a haskell style definition
\begin{lstlisting}
Class Equiv A := (=): relation A.
\end{lstlisting}
Unfortunately, currently this is not supported in Coq.

@@

Although the expression now \emph{looks} like it is bound to some
specific semigroup structure (which with traditional approaches would imply projections from bundles
and/or reference to proofs), it is really only referring near-directly to the actual operations
involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
may use in the proof. This lack of projections keeps our terms small and independent, and keeps
rewriting simple and sane; see Section~\ref{canonical}.

@@

where e.g.\ \lstinline|RingPlus| is the operation class for the semigroup operator.
The notation \lstinline|:>| makes \lstinline|semiring_mult_monoid| a subclass: when asked for a monoid it suffices to
provide a ring.

@@

First, multiple inheritance becomes trivial: \lstinline|SemiRing| inherits two \lstinline|Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.

@@

Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue. Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a \lstinline|SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.

Fourth, \emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
can find the equivalence relation in the context.\footnote{
A similar style should be possible for, say, the \lstinline|Ring| tactic, instead of
declaring the ring structure by a separate command, we would rely on type class resolution to find it.}
We note that \lstinline|op| does not depend on the proof that \lstinline|e| is an equivalence. As explained above we use Coq's implicit quantification (\lstinline|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a theorem. Both features are new in Coq and stem from the type class implementation.
